Introducing region-network priors to statistical estimators for fMRI data

### 2015-11-24 paper/paper.tex
news

\author{Danilo Bzdok, Michael Eickenberg, Olivier Grisel,
  Bertrand Thirion,
  Ga\"el Varoquaux\\
  INRIA, Parietal team, Saclay, France\\
  CEA, Neurospin, Gif-sur-Yvette, France\\
  firstname.lastname@inria.fr}

@@

% OUR keywords
%\textbf{\\keywords}: curse of dimensionality; semi-supervised learning;
%fMRI; systems neuroscience

% official NIPS keywords
\textbf{\\keywords}: Brain Imaging, Functional Specialization,
Functional Integration,
Sparsity-inducing Norms, Systems Biology

@@

\section{Introduction}
%
Many quantitative scientific domains underwent a
recent passage from the classical regime (i.e., "long data")  to
the high-dimensional regime (i.e., "wide data")
\citep{jordan2015massive}.
Also in the brain imaging domain,
many contemporary methods for acquiring brain signals yield
more variables per observation than
total observations available in a data sample.
This scenario challenges many statistical estimators from
classical statistics.
For instance,
linear regression models without any additional assumptions
yield an infinity of possible coefficients
and, thus, no solutions.
%
\textit{Sparsity} assumptions have consequently been
introduced and made many ill-posed estimation problems tractable
\cite{hastie2015statistical}.
Domain-specific structural knowledge can be imposed on the 
statistical estimation by preassuming that not all variables will be
equally important.
Sparsified statistical estimators have proven to yield
statistical relationships that can be
estimated, reproduced, and interpreted
\cite{giraud2014introduction}.
%
Yet, what would be the most pertinent neurobiological structure
that can successfully tackle the curse of dimensionality in
functional neuroimaging research?

abc
in current fMRI,
large-scale integration prevails over regional specialization,
although both notions are inextricably linked (Tononi et al., 1998).

It however remains elusive how brain regions and networks relate to
the organism's repertoire of mental operations.

Both regions and netoworks are
probably mechanistically relevant for emerging behavior.

The organization of the cerebral cortex into distinct modules may be described along several dimensions, most importantly, structure, connectivity and function. This conclusion was mainly derived from concurrent invasive examination of microstructure (histological preparation), connectivity (axonal tracing) and functional properties (single cell recordings) in same animals. 
Regarding the (micro-)structural dimension, early histological investigations into the microscopic heterogeneity of the human cerebral cortex have resulted in several detailed, though partially incongruent, anatomical maps (Brodmann 1909; Vogt and Vogt 1919). More recent advances in structural brain mapping have led to the development of observer-independent probabilistic cytoarchitectonic maps in stereotaxic space (Zilles and Amunts 2010). Moreover, the constantly growing field strength of fMRI scanners permits increasingly fin-grained microstructural recordings of the living human brain (Walters et al. 2007).
Regarding the connectional dimension, each cortical area is assumed to possess a unique set of input and output connections. Axonal connectivity between areas can be revealed by injection of a tracing dye that is transported to interconnected brain regions in animals. Axonal tracing studies in primates suggest that cortical areas tend to be connected hierarchically and reciprocally by feed-forward and feed-backward connections (Maunsell and van Essen 1983). Mapping connectivity patterns of a particular part of the cortex may thus allow identifying cortical areas by demonstrating differences between the connections of neighbouring grey matter locations.

Slightly reframing the concept of structure, connectivity and function as three pillars of brain organization, regional specialization of computational processes (i.e., function) can be conceived as being a consequence of both the local (micro-)structural and global connectional properties, as suggested by research in animals. That is, specialization of a particular function is not regarded as an intrinsic property of a brain region that is independent of its connectivity. Rather, input and output connectivity of an area in combination with the local "infrastructure" provided, e.g., by cyto- and chemoarchitecture, crucially determine what particular functions that area can perform (Passingham et al. 2002). Conversely, each particular cortical module is probably characterized by a unique set of input and output connections, which is supported by statistical analyses of cortical connectivity in the primate (Young 1993) and feline (Scannell et al. 1995) cortex. More generally, cortical modules and connections between those actually reflect functional segregation and functional integration, respectively (Friston 2002). In sum, a cortical module of functional specialization is likely to be defined by the intersection of regionally specific microstructure and connectivity patterns.

SPECIAL

Functional
specialization emphasizes that microanatomically distinguishable brain
regions realize distinct functions (Brodmann 1909; Zeki 1978). Indeed,
each microanatomical cortical area has a unique connectional, and thus
perhaps functional, profile in cats (Scannell et al. 1995), monkeys
(Young et al., 1993), and humans (Passingham et al., 2002).

Consistently, early lesion studies of induced behavioral
impairments in rats challenged the dogmatic localization of brain
function (Franz \& Lashley, 1917). Analogously in humans, even
extensive prefrontal lesions do not necessarily become apparant in
everyday behavior (Mesulam, 1990). An implicit preference for
functional specialization might therefore hinder consensus in
fundamental, still unresolved questions in neuroscience, such as the
relation between cognition and emotion (cf. Pessoa 2010, Overwalle
2011).

locally characteristic neuronal populations

@@

Hebb, D. O. The Organization Of Behaviour (Wiley,
1949).

@@

@@

STUDY

@@

\section{Methods}
%
\paragraph{Data.}
As the currently biggest openly-accessible reference dataset,
we chose resources from the Human Connectome Project (HCP)
\cite{barch2013}.
Neuroimaging task data with labels of ongoing cognitive processes
were drawn from 500
healthy HCP participants (cf. Appendix for details on datasets).
18 HCP tasks 
were selected that are known to elicit reliable neural activity
across participants (Table \ref{table_tasks}).
In sum, the HCP task data incorporated 8650 first-level activity maps
from 18 diverse paradigms administered to 498 participants (2 removed
due to incomplete data).
All maps were resampled to a common $60\times72\times60$ space of
3mm isotropic voxels and gray-matter masked (at least 10\% tissue
probability).
The supervised analyses were thus based on labeled HCP task maps with
79,941 voxels of interest representing z-values in gray matter.

@@

>>>>

\author{Danilo Bzdok, Michael Eickenberg, Olivier Grisel,
  Ga\"el Varoquaux, Bertrand Thirion
  INRIA, Parietal team, Saclay, France\\
  CEA, Neurospin, Gif-sur-Yvette, France\\
  firstname.lastname@inria.fr}

@@

\textbf{\\keywords}:
Brain Imaging, Functional Specialization, Functional Integration,
Sparsity-inducing Norms, Numerical Optimization, Systems Neuroscience

@@

\section{Introduction}
% sparsity
Many quantitative scientific domains underwent a
recent passage from the classical regime (i.e., "long data")  to
the high-dimensional regime (i.e., "wide data")
\citep{jordan2015massive}.
Also in the brain imaging domain,
many contemporary methods for acquiring brain signals yield
more variables per observation than
total observations available in a data sample.
This scenario challenges many statistical estimators from
classical statistics.
For instance,
linear regression models without additional assumptions
yield an infinity of possible coefficients
and, thus, no solution.
%
\textit{Sparsity} assumptions have consequently been
introduced and made many ill-posed estimation problems tractable
\cite{buhlmann2011statistics, hastie2015statistical}.
Domain-specific structural knowledge can be imposed on the 
statistical estimation by preassuming that not all variables will be
equally important
\citep{bach2012optimization}.
Sparsified supervised and unsupervised
learning algorithms have proven to yield
statistical relationships that can be readily
estimated, reproduced, and interpreted
\cite{giraud2014introduction}.
%
Yet, what is the most pertinent neurobiological structure
that can successfully tackle the
\textit{curse of dimensionality} in
functional neuroimaging research?

% specialization & integration
Concepts on human brain organization have long been torn
between the two extremes
\textit{functional specialization} and \textit{functional integration}.
Functional specialization emphasizes that microscopically distinguishable
brain regions probably solve distinct classes of computational processes
\citep{kanwisher2010functional}.
Functional integration, in turn, emphasizes that brain function
is probably enabled by complex connections between these
distinct brain regions.
%
This notion was predominantly derived from
invasive examination of \textit{structure} (i.e., histological preparation),
\textit{connectivity}, (i.e., axonal tracing),
and \textit{functional properties}
(i.e., single cell recordings) in same animals.
Regarding functional segregation into specialized regions,
early histological investigations into the microscopic heterogeneity of
the human cerebral cortex have resulted
in several detailed anatomical maps
\citep{{brodmann1909vergleichende, vogt1919allgemeine}.
Regarding axonal connections,
each such cortical area has been observed
to possess a unique set of incoming and outgoing connections
\citep{passingham2002, young93monkey, scannell95cat}.
%
In particular,
the realization of computational processes (i.e., function)
is probably a consequence of both
local cyto- and chemoarchitectonic infrastructure
and its unique global connectivity profile.
While
cortical modules versus connections between those
reflect 
functional specialization versus functional integration,
\citep{friston2002beyond},
these architectural principles are conceptually inextricable in
the realization of mental operations
\citep{tononi1998complexity}.

method -> type of discovery

SPECIAL
Functional specialization has been investigated in many ways.
%
Single cell recordings and microscopic examination
revealed, for instance, the
specialization in the visual cortex into V1, V2, V3, V3A, and V4
\citep{hubel1962receptive, zeki1978functional}.

example: fusiform gyrus
-> lesion -> does not work anymore

Non-invasive brain imaging with
functional magnetic resonance imaging (fMRI) and
positron emission tomography (PET)
enabled the localization of
sensory, motor, and emotional functions to cortical areas
in the living brain
\citep{fristen1997imaging}.

Radioactive mapping the distribution of
neurotransmitter receptors added yet another
local characteristic of neuronal populations
\citep{zilles2009receptor}.

clustering
\citep{behrens03}

High-thoughput computing enabled an 
ultahigh-resolution 3D model of brain anatomy
at macroscopical to near-cellular scale
\citep{amunts2013bigbrain}.

=> methods: non-overlapping, discrete region units

@@

@@

example: DMN network?

@@

=> methods: overlapping, continuous network units
invasive axonal tracing studies,
ICA,
seed-based analyses of BOLD fluctations or
electrophysiological oscillations,
as well as
effective connectivity models
and
graph-theoretical measurements
\citep{friston03dcm, buzsaki2004neuronal,
beckmann2005, bullmore2009complex, jbabdi2013long}.

analysis and integration of information

% study
The present study attempts a combination of specialization and
integration aspects by means of structured sparsity
for high-dimensional inference.
%

adaptive estimator
biologically plausible

recent advances in high-dimensional statistics enable

@@

importance of brain regions and networks is probably a matter of
degree.
We estimate that degree from the data

help reintegrate the divorced
interpretational streams based on
region- and network-focused neuroscientific investigations.

\section{Methods}
%
\paragraph{Rationale}

we need to inject domain knowledge into
statistical estimations to harness the curse of dimensionality.

should be able to estimate voxel level
while taking into account known supravoxel structure.

%
\paragraph{Data.}
As the currently biggest openly-accessible reference dataset,
we chose resources from the Human Connectome Project (HCP)
\cite{barch2013}.
Neuroimaging task data with labels of ongoing cognitive processes
were drawn from 500
healthy HCP participants (cf. Appendix for details on datasets).
18 HCP tasks 
were selected that are known to elicit reliable neural activity
across participants (Table \ref{table_tasks}).
In sum, the HCP task data incorporated 8650 first-level activity maps
from 18 diverse paradigms administered to 498 participants (2 removed
due to incomplete data).
All maps were resampled to a common $60\times72\times60$ space of
3mm isotropic voxels and gray-matter masked (at least 10\% tissue
probability).
The supervised analyses were thus based on labeled HCP task maps with
79,941 voxels of interest representing z-values in gray matter.

@@

Functional specialization and functional integration
in the human brain are
probably both mechanistically relevant for emerging behavior.
However, there is a scarcity of statistical methods
that concomitantly relate brain regions and networks to
an organism's repertoire of mental operations.
